Stage-Splitting a Modal Language

### 2014-06-20 paper/semantics.tex
finished a rough draft of everything before metatheory

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pval$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain hatted variables. In the example above, 
\verb|(|$\hat {\tt y}$\verb|,7)|.
\item Residuals ($\res$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\val$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear---hatted---in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
This judgment, upon encountering a stage \bbtwo\ subexpression, must necessarily search into it in order to find its stage
\bbone\ subexpressions; we call this process \emph{speculation}. 
Speculation is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.

\TODO next

\TODO prev

\TODO reification

\TODO hold

\TODO contexts

@@

%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%	fn (p : int) => 
%		next{
%			fn (b : int) => 
%				prev{f (next {b}, p)}
%		}	
%\end{lstlisting}
\TODO code written, but listings don't work in footnotes! commented out in source until we get working
}.
Given its stage \bbone{} argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in \langTwo.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

@@

\TODO 
If we change the highlighted rules to something slightly different, then we get
precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
our semantics terminates strictly less often than that of \cite{davies96}.

>>>>

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pval$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain hatted variables. In the example above, 
\verb|(|$\hat {\tt y}$\verb|,7)|.
\item Residuals ($\res$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\val$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear---hatted---in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
and the final future environment is gotten by merging the future environments of the subterms.

When this judgment encounters a $\next$ block (\ref{fig:diaSemanticsNP}), it searches into the block's stage \bbtwo\ content to find any contained stage
\bbone\ subexpressions and evaluate them in place.  
We call this search process \emph{speculation}. It is implemented by the $\redtwosym$ judgment, which takes a stage \bbtwo\ term to a residual.
Once the contents of the $\next$ block are speculated into a residual ($q$), 
we then return a fresh hatted variable ($\hat y$), 
along with a future environment which maps that variable to our residual ($y \mapsto q$).

At all of the normal features (\ref{fig:diaSemanticsSpec}), speculation does nothing but recursively speculate into every subexpression.
Once speculation finds a $\prev$ block, we resume stage \bbone\ evaluation of the contents, which produces a future context and (by canonical forms) a hatted variable ($\hat y$).
The context is then reified (using the $\reifysym$ judgment) into a series of let bindings, with $y$---stripped of its hat---at the bottom.

Because the inputs to evaluation and speculation are sometimes open on variables, we adorn our judgments with two contexts.
The first of these ($\Xi$) keeps track of those hatted variables that may appear in the input.
Since hatted variables can only be introduced to terms via substitution ({\em i.e.} they do not appear in the original program), 
this context is augmented only at stage \bbone\ substitution sites.
The second context ($\Gamma$) keeps track of stage \bbtwo\ variables, which are declared in original program.
% We chose to keep these as two separate contexts, because only $\Xi$ is needed for well-formedness of partial values.

If we change the $\next$ and $\prev$ rules to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value, where $q$ is a residual,
then we get precisely the semantics from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
But moreover, because a reified residual will always be evaluated whether or not its result is consumed, 
our semantics terminates strictly less often than that of \cite{davies96}.

\subsection {A Partial Evaluation System}

@@

%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%	fn (p : int) => 
%		next{
%			fn (b : int) => 
%				prev{f (next {b}, p)}
%		}	
%\end{lstlisting}
\TODO code written, but listings don't work in footnotes! commented out in source until we get working
}.
Given its stage \bbone\ argument $a:A$, we can evaluate the partially-applied
function:
$\cdot;\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$. Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $v$, via $\reify\xi{v}{f_a}$. A canonical forms theorem on $v$
ensures that the resulting $f_a$ has type $B\to C$ in the residual language.
Finally, given a stage \bbtwo{} argument $b:B$, we can stage-\bbtwo{} evaluate
the ultimate result of the function, $f_a~b \mathop{\tworedsym} c$.

@@
